Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-08-03 bmc_template/figures.Rnw
everything before supplementary is done

<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Simulation test for BIC-optimal approach",fig.width=6,fig.height=4>>=
ICdat <- readRDS("../sim_results/ICdat.rds")
aa <- readRDS("../sim_results/sim.101.RDS")

ICdat <- mutate(ICdat,
                model=factor(model,
                         levels=c("HMM","HMMsin"),
                         labels=c("homogeneous","heterogeneous"))) %>% filter(IC=="BIC")
simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model,
                            linetype=IC,shape=IC))+
                        geom_point(size=3) +
                        geom_line() +
                        scale_colour_brewer(palette="Set1")+
                        labs(y="counts",x="# latent states")
print(simplot)

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.width=6,fig.height=4>>=
allmodsumdf2 <- readRDS("../summary_stats/cat.1.df.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"time-geneous HMM","homogeneous HMM"))
(ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))
    + geom_line()
    + facet_wrap(~ homog)
    + scale_colour_brewer(name="step length\ndistribution",palette="Set1")
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    + scale_linetype(name="von Mises\nturning angle?")
    + labs(y=expression(Delta*"BIC"),
         x="Number of States")
    + geom_hline(yintercept=0,colour="gray")
    + zmargin)
@

@@

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,,fig.width=8>>=
#L <- load("cat1sumdat.RData")

@@

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE,fig.width=8>>=

@@

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE>>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
vitacf <- (vit %>% dplyr::select(-c(time,obs))%>%ldply(.,acffun) %>% mutate(model=.id)
)

@@

<<r msdlist,message=FALSE, echo=FALSE, warning=FALSE>>=
get_cat_pars2 <- function(catid){
  templist <- readRDS(paste("../summary_stats/cat.",catid,".df.RDS",sep=""))
  moddf <- templist$msdlist

@@

\section{sup}

>>>>

<<sim_results, warning=FALSE,echo=FALSE,message=FALSE,fig.cap="BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.",fig.width=6,fig.height=4>>=
ICdat <- readRDS("../sim_results/ICdat.rds")
aa <- readRDS("../sim_results/sim.101.RDS")

ICdat <- mutate(ICdat,
                model=factor(model,
                         levels=c("HMM","HMMsin"),
                         labels=c("Homogeneous","Heterogeneous"))) %>% filter(IC=="BIC")
simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model))+
                        geom_point(size=3) +
                        geom_line() +
                        scale_colour_brewer(palette="Set1")+
                        labs(y="Counts",x="BIC-optimal States")
print(simplot)

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.",fig.width=6,fig.height=4>>=
allmodsumdf2 <- readRDS("../summary_stats/cat.1.df.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"Time-inhomogeneous HMM","Homogeneous HMM"))
(ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))
    + geom_line()
    + facet_wrap(~ homog)
    + scale_colour_brewer(name="Step-length\ndistribution",palette="Set1")
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    + scale_linetype(name="von Mises\nturning angle?")
    + labs(y=expression(Delta*"BIC"),
         x="Number of States")
    + geom_hline(yintercept=0,colour="gray")
    + zmargin)
@

@@

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Adjusted BIC by number of free parameters for HMM model types. The left panel shows FMM, FMM with a sin prior and HMM. The right panel shows HMMs with different temporal transitions.",fig.width=8>>=
#L <- load("cat1sumdat.RData")

@@

<<avg_step_length_by_time, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.",fig.width=8>>=

@@

<<acf_plot, message=FALSE, echo=FALSE, warning=FALSE,fig.cap="Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states ">>=
acffun <- function(sim){
  a <- acf(sim,na.action=na.pass,plot=FALSE)
  df <- data.frame(ACF=c(a$acf),lag=a$lag)
  return(df)
}
vitacf <- (vit %>% dplyr::select(-c(time,obs))%>%ldply(.,acffun) %>% mutate(model=.id)
)

@@

<<r msdlist,message=FALSE, echo=FALSE, fig.cap="State identification plot with log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.", warning=FALSE>>=
get_cat_pars2 <- function(catid){
  templist <- readRDS(paste("../summary_stats/cat.",catid,".df.RDS",sep=""))
  moddf <- templist$msdlist

@@

\section{Additional files}

### 2016-08-03 bmc_template/figures.tex
everything before supplementary is done

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/sim_results-1} \caption[Simulation test for BIC-optimal approach]{Simulation test for BIC-optimal approach}\label{fig:sim_results}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/BICred_plot-1}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons-1}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time-1}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/acf_plot-1}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/r_msdlist-1}

>>>>

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/sim_results-1} \caption[BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations]{BIC-optimal state frequency for 2-6 state HMMs with and without covariate transition on 100 two-state hidden Markov models with covariate transitioning simulations.}\label{fig:sim_results}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/BICred_plot-1} \caption[Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles]{Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.}\label{fig:BICred_plot}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/adj_BIC_comparisons-1} \caption[Adjusted BIC by number of free parameters for HMM model types]{Adjusted BIC by number of free parameters for HMM model types. The left panel shows FMM, FMM with a sin prior and HMM. The right panel shows HMMs with different temporal transitions.}\label{fig:adj_BIC_comparisons}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/avg_step_length_by_time-1} \caption[Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states]{Average step length by time of day observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states.}\label{fig:avg_step_length_by_time}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/acf_plot-1} \caption[Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states ]{Autocorrelation of observed (gray highlight), three-state HMM Viterbi predictions (V points), and all transitions type HMMs predictions (out of sample) with their respective BIC-optimal states }\label{fig:acf_plot}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/r_msdlist-1} \caption[State identification plot with log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition]{State identification plot with log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.}\label{fig:r msdlist}
\end{figure}

### 2016-08-03 bmc_template/manuscript.tex
everything before supplementary is done

\

@@

\

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transition probabilities -- in particular,
transition probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  
In addition to simulation-based examples,
we also re-analyze data from 
van de Kerk et al.\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
describe the movement and putative underlying
behavioural states of Florida panthers (\emph{Puma concolor coryi}).
% 
% %%\todo{we will clarify below that they actually used HSMMs} 
% %%to analyze hourly movement data
% %%from 18 Florida panthers (\emph{Puma concolor coryi}).
% 
\

van de Kerk et al.\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as six for individuals with a large amount
of available data); for reasons of computational 
practicality and biological interpretability, they restricted their
detailed analysis to models with only three underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\cite{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\cite{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
In contrast, \cite{dean_behavioural_2013} evaluated models with up to 10
states, but like \cite{kerk2015hidden} they chose to consider only 
models with three states. 
% (their criterion for choosing ...\todo{check \cite{zhao_knee_2008}!}).
As van de Kerk et al. \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one possible reason that other authors have not adopted
van de Kerk et al.'s model-based approach to
identifying the number of latent states.

\

@@

\

van de Kerk et al.\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
three to six states,
suggesting that the six-state model was
favoured statistically; however,
the authors used three-state models in most
of their analyses for ease of biological interpretation.
We follow van de Kerk et al.\cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.  
% \todo{it occurs to me that we haven't really included any
% information about simulation testing in this version!
% We should probably add a short section; if don't, then we
% should delete following sentence }
Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\cite{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \cite{BurnhamAnderson1998};
we believe that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

\

@@

\

\begin{description}
\item[Multiple block transition] Here we assume piecewise-constant transition probabilities. The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:

$$
\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm} \delta_{m=t}
$$
where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block at the corresponding time $t$, and 0 otherwise).

\

\item[Quadratic transition model] We assume the elements of the linear predictor
are quadratic functions of hour. The quadratic model is not diurnally continuous, i.e. there is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity constraint would collapse the model to a constant.
$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}\left(\frac{t}{24}\right)+b_{ij3}\left(\frac{t}{24}\right)^{2}
$$

\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint.
$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
$$

@@

\

\subsection{Model evaluation}

\

@@

\

\section{Results}

\

We simulated a two-state HMM with sinusoidal temporal transitions 100 times and fitted it with two to five state HMMs and without temporal transition.  Fig. ~\ref{fig:sim_results} shows heterogeneous transition models can always predict the correct number of states, whereas, can overestimate the number of states via BIC-optimal approach.

\

Fig. ~\ref{fig:BICred_plot} shows that the BIC-optimal number of states for time homogenous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles)

\

@@

\

Figure~\ref{fig:adj_BIC_comparisons} shows that models with temporal heterogenity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogenous HMMs are better than FMMs with sinusoidal temporal heterogeneity. Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states. The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
However, this similarity is somewhat overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
normally be interpreted as an enormous difference in goodness of fit,
between the sinusoidal and quadratic models (both of which have 90 
parameters).

\

Fig. ~\ref{fig:avg_step_length_by_time} shows a clear diurnal pattern from the average hourly step lengths from the observed data. As expected, temporally homogeneous models (whether FMM or HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle. All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns. However, the block model does markedly worse than the other temporal models (changing the block definitions might help), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough). We also included average hourly step lengths from three-state temporally homogeneous HMM Viterbi prediction (v points).

\

Like the diurnal pattern shown in Fig. ~\ref{fig:avg_step_length_by_time},
the strong autocorrelation of the observed step lengths at a 24-hour lag (Fig. ~\ref{fig:acf_plot}) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
Because there are a huge number of potential complexities that can be added to movement models (e.g. spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable. In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects. HMM without temporal heterogeneity captures the short-term autocorrelation, but misses the long-term autocorrelation beyond a 7-hour lag.
Temporally homogeneous FMM, by definition, produces neither short- nor long-term autocorrelation. FMM without temporal heterogeneity, although it captures the diurnal pattern well, underpredicts the degree of short-term autocorrelation.

\

The hardest problem with multiple latent states is interpreting them
biologically. We have no way of knowing what panthers are actually
thinking (it is certainly more complex than being in one of a small
number of discrete latent states); 
we don't know the ``true'' number of latent states,
nor are we able to observe them directly, although
incorporating additional direct observations of 
behaviour (if available) can at least partially
address this problem \cite{fryxell_multiple_2008}.
Three distinct
movement states seem biologically interpretable for Florida panthers
according to van de Kerk et al.\cite{kerk2015hidden}: Short step length suggests resting
states, intermediate step length a foraging state, and long
step length a traveling state. 
% \todo{do we need this?}
Figure~\ref{fig:state_plots} compares the estimated parameter values
for several cats
(mean and standard deviation of the step length in each state) between
the time-homogeneous and time-heterogeneous models. 
In general, the states with longer mean step lengths are relatively
similar between model classes. For cats 14 and 15, the states with
the longest or next-longest mean step lengths have similar means
and standard deviations; for cats 1 and 2, three long-step states
in the homogeneous HMM appear to divide two long-step states in 
the heterogeneous HMM. For short-step states, the heterogeneous HMM
tends to identify a high-variance state, while the homogeneous HMM
picks up states with very short step lengths (questionable in any
case because we have not taken any special efforts to account for
GPS error).

\

\section{Discussion}

\

@@

\

\section{Conclusion}

\

@@

\section{Declarations}

@@

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \begin{figure}[h!]
  \includegraphics[width=\maxwidth]{figure/sim_results-1}
  \caption{\csentence{Sample figure title.}
      A short description of the figure content
      should go here.}
      \end{figure}

\begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      Figure legend text.}
      \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Sample table title. This is where the description of the table should go.}
      \begin{tabular}{cccc}
        \hline
           & B1  &B2   & B3\\ \hline
        A1 & 0.1 & 0.2 & 0.3\\
        A2 & ... & ..  & .\\
        A3 & ..  & .   & .\\ \hline
      \end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

>>>>

@@

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transition probabilities -- in particular,
transition probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  
In addition to simulation-based examples,
we also re-analyze data from 
van de Kerk et al.\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
describe the movement and putative underlying
behavioural states of Florida panthers (\emph{Puma concolor coryi}).

van de Kerk et al.\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as six for individuals with a large amount
of available data); for reasons of computational 
practicality and biological interpretability, they restricted their
detailed analysis to models with only three underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\cite{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\cite{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
In contrast, \cite{dean_behavioural_2013} evaluated models with up to 10
states, but like \cite{kerk2015hidden} they chose to consider only 
models with three states. 
As van de Kerk et al. \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one possible reason that other authors have not adopted
van de Kerk et al.'s model-based approach to
identifying the number of latent states.

@@

van de Kerk et al.\cite{kerk2015hidden} used the Bayesian (Schwarz) information criterion (BIC)
to test the relative penalized goodness of fit for
models ranging from 2 to 6 latent states.
In general, BIC values decreased as the number of states
increased from 
three to six states,
suggesting that the six-state model was
favoured statistically; however,
the authors used three-state models in most
of their analyses for ease of biological interpretation.
We follow van de Kerk et al.\cite{kerk2015hidden} in using BIC-optimality (i.e., minimum
BIC across a family of models) as the criterion for identifying
the best model, because we are interested in explaining the 
data generation process by identifying the ``true'' number 
of underlying movement states.

Using BIC also simplifies 
evaluation of model selection procedures; 
it is easier to test whether our model selection
procedure has selected the model used to simulate the data,
rather than testing whether it has selected the model with
the minimal Kullback-Leibler distance
\cite{Richards2005}. We recognize that 
ecologists will often be interested in maximizing predictive
accuracy rather than selecting a true model, and that as
usual in ecological systems the true model will be far more
complicated than any candidate model \cite{BurnhamAnderson1998};
we believe that the qualitative conclusions stated here
for BIC-optimality will carry over to analyses using AIC instead.

@@

\begin{description}

\item[Multiple block transition] Here we assume piecewise-constant transition probabilities. The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:

$$
\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm} \delta_{m=t}
$$

where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block at the corresponding time $t$, and 0 otherwise).

\item[Quadratic transition model] 
We assume the elements of the linear predictor
are quadratic functions of hour. The quadratic model is not diurnally continuous, i.e. there 
is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity 
constraint would collapse the model to a constant.

$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}\left(\frac{t}{24}\right)+b_{ij3}\left(\frac{t}{24}\right)^{2}
$$

\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint.

$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
$$

@@

\subsection*{Model evaluation}

@@

\section*{Results}

%% fig 1 goes somewhere here

We simulated a two-state HMM with sinusoidal temporal transitions 100 times and fitted it with two to five state HMMs and without temporal transition. Heterogeneous transition models can always predict the correct number of states, whereas, can overestimate the number of states via BIC-optimal approach (Figure 1).

%% fig 2 goes somewhere here

The BIC-optimal number of states for time homogenous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results (Weibull wrapped-Cauchy to Weibull von Mises, and Weibull von Mises to log Normal without turning angles; Figure 2)

@@

%% fig 3 goes somewhere here

Models with temporal heterogenity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogenous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3). Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states. The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
However, this similarity is somewhat overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
normally be interpreted as an enormous difference in goodness of fit,
between the sinusoidal and quadratic models (both of which have 90 
parameters).

The panthers exhibits a clear diurnal pattern from the average hourly step lengths from the observed data (Figure 4). As expected, temporally homogeneous models (whether FMM or HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle. All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns. However, the block model does markedly worse than the other temporal models (changing the block definitions might help), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough). We also included average hourly step lengths from three-state temporally homogeneous HMM Viterbi prediction (v points).

Like the diurnal pattern (Figure 4),
the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure 5) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
Because there are a huge number of potential complexities that can be added to movement models (e.g. spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable. In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects. HMM without temporal heterogeneity captures the short-term autocorrelation, but misses the long-term autocorrelation beyond a 7-hour lag.
Temporally homogeneous FMM, by definition, produces neither short- nor long-term autocorrelation. FMM without temporal heterogeneity, although it captures the diurnal pattern well, underpredicts the degree of short-term autocorrelation.

The hardest problem with multiple latent states is interpreting them
biologically. We have no way of knowing what panthers are actually
thinking (it is certainly more complex than being in one of a small
number of discrete latent states); 
we don't know the ``true'' number of latent states,
nor are we able to observe them directly, although
incorporating additional direct observations of 
behaviour (if available) can at least partially
address this problem \cite{fryxell_multiple_2008}.
Three distinct
movement states seem biologically interpretable for Florida panthers
according to van de Kerk et al.\cite{kerk2015hidden}: Short step length suggests resting
states, intermediate step length a foraging state, and long
step length a traveling state.

The estimated parameter values
for several cats
(mean and standard deviation of the step length in each state) between
the time-homogeneous and time-heterogeneous models are similar across all cats (Figure 7). 
In general, the states with longer mean step lengths are relatively
similar between model classes. For cats 14 and 15, the states with
the longest or next-longest mean step lengths have similar means
and standard deviations; for cats 1 and 2, three long-step states
in the homogeneous HMM appear to divide two long-step states in 
the heterogeneous HMM. For short-step states, the heterogeneous HMM
tends to identify a high-variance state, while the homogeneous HMM
picks up states with very short step lengths (questionable in any
case because we have not taken any special efforts to account for
GPS error).

\section*{Discussion}

@@

\section*{Conclusion}

@@

\section*{Declarations}

@@
